---
title: <span style="color:#0085A1; font-size:5rem">Generalized Additive Models</span>
author: |
  | <a href="http://mclark--.github.io/"><span style="font-size:2rem">Michael Clark</span></a>
  | <span style="color:#00274c">Statistician Lead
  | Consulting for Statistics, Computing and Analytics Research
  | Advanced Research Computing </span>
date: '`r Sys.Date()`'
output:
  html_document:
    css: tufte-css-master/tufte.css
    highlight: pygments
    keep_md: no
    theme: cosmo
    toc: yes
    toc_float: yes
bibliography: refs.bib
nocite: | 
  @wood_generalized_2006, @venables_modern_2002, @rasmussen_gaussian_2006, @hardin_generalized_2012, 
  @rigby_generalized_2005, @hastie_generalized_1990, @fox_multiple_2000, @fox_nonparametric_2000,
  @breiman_statistical_2001, @bybee_pisa_2009, @hastie_elements_2009, @ruppert_semiparametric_2003,
  @wasserman_all_2006, @fahrmeir2013regression, @friedman2000additive
---
#<a name='appendix'>Appendix</a>

##<a name='packages'>R packages</a>
<span class="newthought">The following is</span> a non-exhaustive list of R packages which contain GAM functionality. Each is linked to the CRAN page for the package.  Note also that several build upon the <span class="pack">mgcv</span> package used for this document.

[brms](http://cran.r-project.org/web/packages/brms/) Allows for Bayesian GAMs via the Stan modeling language (very new implementation).

[CausalGAM](http://cran.r-project.org/web/packages/CausalGAM/) This package implements various estimators for average treatment effects. 

[COZIGAM](http://cran.r-project.org/web/packages/COZIGAM/) Constrained and Unconstrained Zero-Inflated Generalized Additive Models.

[CoxBoost](http://cran.r-project.org/web/packages/CoxBoost/) This package provides routines for fitting Cox models. See also <span class="func">cph</span> in rms package for nonlinear approaches in the survival context.

[gam](http://cran.r-project.org/web/packages/gam/) Functions for fitting and working with generalized additive models.  

[GAMBoost](http://cran.r-project.org/web/packages/GAMBoost/): This package provides routines for fitting generalized linear and and generalized additive models by likelihood based boosting.

[gamboostLSS](http://cran.r-project.org/web/packages/gamboostLSS/): Boosting models for fitting generalized additive models for location, shape and scale (gamLSS models). 

[GAMens](http://cran.r-project.org/web/packages/GAMens/): This package implements the GAMbag, GAMrsm and GAMens ensemble classifiers for binary classification.

[gamlss](http://cran.r-project.org/web/packages/gamlss/): Generalized additive models for location, shape, and scale. 

[gamm4](http://cran.r-project.org/web/packages/gamm4/): Fit generalized additive mixed models via a version of mgcv's gamm function.  

[gammSlice](http://cran.r-project.org/web/packages/gammSlice/): Bayesian fitting and inference for generalized additive mixed models.

[GMMBoost](http://cran.r-project.org/web/packages/GMMBoost/): Likelihood-based Boosting for Generalized mixed models.

[gss](http://cran.r-project.org/web/packages/gss/):  A comprehensive package for structural multivariate function estimation using smoothing splines.

[mboost](http://cran.r-project.org/web/packages/mboost/): Model-Based Boosting. 

[mgcv](http://cran.r-project.org/web/packages/mgcv/): Routines for GAMs and other generalized ridge regression with multiple smoothing parameter selection by GCV, REML or UBRE/AIC. Also GAMMs. 

[VGAM](http://cran.r-project.org/web/packages/VGAM/): Vector generalized linear and additive models, and associated models.



##<a name='Penalized_Estimation_Example'>Penalized Estimation Example</a>

Initial data set up and functions.

```{r Penalized_Estimation_Example}
############################
### Wood by-hand example ###
############################


size = c(1.42,1.58,1.78,1.99,1.99,1.99,2.13,2.13,2.13,
         2.32,2.32,2.32,2.32,2.32,2.43,2.43,2.78,2.98,2.98)
wear = c(4.0,4.2,2.5,2.6,2.8,2.4,3.2,2.4,2.6,4.8,2.9,
         3.8,3.0,2.7,3.1,3.3,3.0,2.8,1.7)
x= size-min(size); x = x/max(x)
d = data.frame(wear, x)

#cubic spline function
rk <- function(x,z) {
  ((z-0.5)^2 - 1/12)*((x-0.5)^2 - 1/12)/4-
    ((abs(x-z)-0.5)^4-(abs(x-z)-0.5)^2/2 + 7/240) / 24
}

spl.X <- function(x,knots){
  q <- length(knots) + 2           # number of parameters
  n <- length(x)                   # number of observations
  X <- matrix(1,n,q)               # initialized model matrix
  X[,2] <- x                       # set second column to x
  X[,3:q] <- outer(x,knots,FUN=rk) # remaining to cubic spline
  X
}

spl.S <- function(knots) {
  q = length(knots) + 2
  S = matrix(0,q,q)          # initialize matrix
  S[3:q,3:q] = outer(knots,knots,FUN=rk) # fill in non-zero part
  S
}

#matrix square root function
mat.sqrt <- function(S){
  d = eigen(S, symmetric=T)
  rS = d$vectors%*%diag(d$values^.5)%*%t(d$vectors)
  rS
}

#the fitting function
prs.fit <- function(y,x,knots,lambda){
  q = length(knots) + 2      # dimension of basis
  n = length(x)              # number of observations
  Xa = rbind(spl.X(x,knots), mat.sqrt(spl.S(knots))*sqrt(lambda))  # augmented model matrix
  y[(n+1):(n+q)] = 0         #augment the data vector
  lm(y ~ Xa-1)               # fit and return penalized regression spline
}
```

Example 1.

```{r Penalized_Estimation_Example_Example1, dev.args = list(bg = 'transparent')}
knots = 1:4/5
X = spl.X(x,knots)           # generate model matrix
mod.1 = lm(wear~X-1)         # fit model
xp <- 0:100/100              # x values for prediction
Xp <- spl.X(xp, knots)       # prediction matrix

# ggplot
library(ggplot2)
ggplot(aes(x=x, y=wear), data=data.frame(x,wear))+
  geom_point(color="#FF8000") +
  geom_line(aes(x=xp, y=Xp%*%coef(mod.1)), data=data.frame(xp,Xp), color="#2957FF") +
  theme_trueMinimal() +
  theme(plot.background=element_blank(),
        panel.background=element_blank())
```


Example 2.


```{r Penalized_Estimation_Example_Example12, dev.args = list(bg = 'transparent')}
knots = 1:7/8

d2 = data.frame(x=xp)

for (i in c(.1,.01,.001,.0001,.00001,.000001)){
  mod.2 = prs.fit(wear, x, knots,i)   #fit penalized regression 
                                      #spline choosing lambda
  Xp = spl.X(xp,knots) #matrix to map parameters to fitted values at xp
  d2[,paste('lambda = ',i, sep="")] = Xp%*%coef(mod.2)
}

### ggplot
library(ggplot2); library(tidyr)
d3 = gather(d2, key=variable, value=value, factor_key=T, -x)

ggplot(aes(x=x, y=wear), data=d) +
  geom_point(col='#FF8000') +
  geom_line(aes(x=x,y=value), col="#2957FF", data=d3) +
  facet_wrap(~variable) +
  theme_trueMinimal() +
  theme(plot.background=element_blank(),
        panel.background=element_blank())
```

