---
title: <span style="color:#0085A1; font-size:5rem">Generalized Additive Models</span>
author: |
  | <a href="http://mclark--.github.io/"><span style="font-size:2rem">Michael Clark</span></a>
  | <span style="color:#00274c">Statistician Lead
  | Consulting for Statistics, Computing and Analytics Research
  | Advanced Research Computing </span>
date: '`r Sys.Date()`'
output:
  html_document:
    css: tufte-css-master/tufte.css
    highlight: pygments
    keep_md: no
    theme: cosmo
    toc: yes
    toc_float: yes
    # code_folding: show
bibliography: refs.bib
nocite: | 
  @wood_generalized_2006,@venables_modern_2002
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = T, message = F, warning = F, R.options=list(width=120), 
                      fig.height=4, fig.width=4, fig.align = 'center', comment=NA, autodep=T)
library(magrittr); library(dplyr); library(pander); library(plotly); library(lazerhawk)
```


# [Preface](#preface)

The following provides a brief introduction to generalized additive models and some thoughts on getting started within the R environment.  It doesn't assume much more than a basic exposure to regression, and maybe a general idea of R though not necessarily any particular expertise. The presentation is of a very applied nature, and such that the topics build upon the familiar and generalize to the less so, with the hope that one can bring the concepts they are comfortable with to the new material. The audience in mind is a researcher with typical social science training, though is not specific to those disciplines alone.

As this document is more conceptual, a basic familiarity with R is all that is needed to follow the code, though there is much to be gained from simple web browsing on R if one needs it. And while it wasn't the intention starting out, this document could be seen as a vignette for the <span class="pack">mgcv</span> package, which is highly recommended.


This document was created within [Rstudio](http://rstudio.org/) and rmarkdown. <span class="marginnote">Last modified `r Sys.Date()`. Original draft August, 2012.</span>


Color guide:

- <span class="emph">important term</span>
- [link]()
- <span class="pack">package</span>
- <span class="func">function</span>
- <span class="objclass">object or class</span>

R Info: **`r sessionInfo()$R.version$version.string` `r sessionInfo()$R.version$nickname`**



```{r child='introduction.Rmd'}
```

```{r child='application.Rmd'}
```


# References
